---
title: "purrr Tricks with All Subset Regression"
author: "Mark Ewing"
description: "See how this one clickbaity purrr trick helped me solve an issue with all subsets regression!"
date: 2017-08-24
categories: ["R","Functional Programming"]
tags: ["rstudio"]
type: "post"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(magrittr)
library(leaps)
library(purrr)
library(stringr)
library(dplyr)
```

## All Subsets Regression

What is all subsets regression?  It's a technique for model building which involves taking a set of independent variables $X_1..i$ and regressing them in sets of $k$, where $k$ is in $\{1,2,\dots,i\}$, against the response variable $Y$.  The 'all' part of 'all subsets' means it's every combination of $X_{1..i}$ being drawn $k$ at a time.

### But...why?

You're probably familiar with forward, backward or stepwise model building where terms are added (or removed) from a model one at a time while attempting to maximize or minimize some 'goodness' criteria.  These are generally regarded as bad - so why all subsets?  Really, it has all the faults of the other methods but it has at least one advantage - you can have more complex 'goodness' criteria which may hit local maxima/minima (interupting the previous methods) but allow you to find the global.  So, it's fishing to the max.

### leaps

R has a great package called `leaps` which implements all subsets regression.  But...it doesn't meet my needs.

## All Subsets Regression and Missing Values

Let's consider a dataset which has missing values.  I'll generate a dummy set of data containing 4 independent variables and one dependent which depends on the first three variables.  Then, I'll randomly assign ~12% of the data to be missing.

```{r}
set.seed(1001)
X1 = c(-1,0,1)
df = expand.grid(X1=X1,X2=X1,X3=X1,X4=X1)
df$Y = df$X1+3*df$X2-0.5*df$X3+rnorm(81,0,1)
df$X1[sample(1:81,10,replace = FALSE)] = NA
df$X2[sample(1:81,10,replace = FALSE)] = NA
df$X3[sample(1:81,10,replace = FALSE)] = NA
df$X4[sample(1:81,10,replace = FALSE)] = NA
```

We can verify that no rows are completely missing:

```{r}
any(apply(is.na(df[,1:4]),1,all))
# Explanation
# df %>% 
#   #only consider the first 4 columns
#   .[,1:4] %>% 
#   #convert to a logical data frame checking for missing values
#   is.na() %>% 
#   #this function applies another function row or column wise to the input dataframe or matrix
#   apply(
#     # 1 means rows
#     1, 
#     # Check if all the values are TRUE
#     all) %>% 
#   # now we have a vector of logicals checking if the whole row of df was missing
#   # check if any are TRUE
#   any()
```

And we can count that there are 50 complete rows (just 62% of the original data).

```{r}
sum(apply(!is.na(df[,1:4]),1,all))
# Explanation
# df %>% 
#   #only consider the first 4 columns
#   .[,1:4] %>% 
#    #convert to a logical data frame checking for missing values
#   is.na() %>%
#   #flip the logical values
#   `!` %>% 
#   #this function applies another function row or column wise to the input dataframe or matrix
#   apply(
#     # 1 means rows
#     1,
#     # Check if all the values are TRUE
#     all
#   ) %>% 
#   # now we have a vector of logicals checking if the whole row of df is present
#   # sum this up to get a count
#   sum()
```

In linear regression observations cannot be included if they are missing because we don't know what value they take.  And sure, there's missing value imputation - as always in statistics, if you're willing to make assumptions whatever you're doing becomes more statistically powerful, but also more real life fragile.  With missing value imputation you have to make some assumptions about why they're missing and then you have to generate the values themselves which requires another set of assumptions.  It's fine to not want to make those assumptions.

So despite the fact that we have 81 rows, we only have 50 rows for fitting a linear regression model which means if we fit the full model we'll only have $n-1-k=50-1-4=45$ degrees of freedom.

```{r}
fullModel = lm(Y~X1+X2+X3+X4,data=df)
summary(fullModel)
```

You can see the DF agree with what we expect to see.  Also, please ignore the fact that we're getting results which are nearly perfect.  It's hard to generate dummy data that looks like real world data.

### First Subset

The first subset is typically just the first indpendent variable we encounter, on it's own, so X1 in our case, but we'll start with X1+X2 (because it helps with comparing to leaps).  We have 61 complete observations across X1 and X2.

```{r}
sum(apply(!is.na(df[,1:2]),1,all))
```

So when we fit the model we expect to have $n-1-k=61-1-2=58$ degrees of freedom.

```{r}
firstSubset = lm(Y~X1+X2,data=df)
summary(firstSubset)
```

This is a pretty good model, and we nailed basic arithmetic!  How does this model compare to what we would get if we used only the complete observations?

```{r}
firstSubset_limited = lm(Y~X1+X2,data=fullModel$model)
summary(firstSubset_limited)
```

It's not very different, but the R^2 value is a little higher.  Also the AIC and BIC are little better for the reduced data set.

```{r}
cat("All available observations AIC:    ",AIC(firstSubset),"\nJust the complete observations AIC:",AIC(firstSubset_limited),"\n\nAll available observations BIC:    ",BIC(firstSubset),"\nJust the complete observations BIC:",BIC(firstSubset_limited))
```

### leaps

So what does leaps produce?  The `regsubsets` lets us specify a formula just like in lm and then it fits all subsets of the specified full model.  The summary shows the best model for each set size $k$.  It selected X2 in the $k=1$ case, X1+X2 in the $k=2$ case, X1+X2+X3 in the $k=3$ and the full model for $k=4$.

```{r}
library(leaps) #Version 3.0
leapsResults = leaps::regsubsets(Y~X1+X2+X3+X4,data=df)
summary(leapsResults)
```

The summary object has a lot in it including BIC, $R^2_{Adj}$, Mallows Cp and more.  It's important to note that `leaps` only retains information about the best models selected.

It looks like the way `stats::BIC` calculates BIC and the way `leaps` does don't match up.  That won't be much use.  But look at the second value of $R^2_{Adj} = 0.7726$ - that's what we calculated when using only the complete observations across all variables.

```{r}
leapsSummary = summary(leapsResults)
leapsSummary$bic[2] #second value for the two term model
leapsSummary$adjr2[2]
```

## How much data to use?

Using only the complete observations in all subset regression is common across several platforms including R and JMP.  But why? Is it more statistically valid to ignore data that could be part of the regression model? Or is it simply more computationally efficient?  It seems to me that each model you fit should be fit using all data available to it - why should my results be different because I collected results on X3 and X4?  I should get the exact same results when fitting the selected model as when doing my model building.  So, how can we do all subsets regression using all the data available?  And when do we get to purrr?

## New Functions

Below I present 3 functions with a limited amount of safe programming tossed in.  

### allSubsetsRegression

The main function is `allSubsetsRegression` which is too long of a name.  It accepts 
* a data frame which should contain only the independent variables to be included and the dependent variable
* a quoted string for the column name in `data` which is the independent variable
* optionally, minimum number of variables to include
* optionally, maximum number of variables to include (useful if your data is wider than it is tall)

This makes a call to `calcKreg`

```{r}
allSubsetsRegression = function(data,y,minV=1L,maxV){
  if(missing(maxV)) maxV = as.integer(ncol(data)-1)
  if(minV > maxV) stop("Max number of variables must be bigger than the min number of variables")
  if(!is.integer(minV) | !is.integer(maxV)) stop("Min/Max variables must be integers")
  if(!is.data.frame(data)) stop("Data must be provided in a data frame")
  if(!is.character(y)) stop("Response variable 'y' must be provided as a quoted string")
  xnames = names(data)[names(data) != y]
  output = lapply(minV:maxV,calcKreg,data=data,y=y,xnames=xnames)
  class(output) = c(class(output),"subReg")
  return(output)
}
```

### calcKreg

The name looks like "calc Kreg" which sounds like a weird German thing but it's actually "calc K reg" meaning "calculate the $k^{th}$ regression set".  This requires
* the provided data
* the quoted string of the dependent variable
* the vector of quoted strings of the independent variables
* k - how big the set is

This leverages the built in function `combn` which returns a matrix of all combinations of $X_i$ of size $k$.  This in turn calls the function `genReg`.

```{r}
calcKreg = function(data,y,xnames,k){
  runs = combn(xnames, k)
  models = lapply(1:ncol(runs), genReg, data=data, y=y, runs=runs)
  names(models) = apply(runs, 2, paste, collapse='+')
  return(models)
}
```

### genReg

`genReg` ("generate regression") requires
* the provided data
* the quoted string of the dependent variable
* the output of `combn`
* the column being used this run

It generates a formula by pasting the needed independent variables together with "+" and then returns the full lm object.

```{r}
genReg = function(data,y,runs,i){
  x = runs[,i]
  form = paste0(y,"~",paste(x,collapse="+"))
  return(lm(form,data=data))
}
```

#### DANGER, WILL ROBINSON! DANGER!

I'm returning the *entire* lm object?  Well, yeah.  If you have stupid big data or stupid wide data, this won't work well, but also, you probably aren't really wanting to do linear regression.  I'm just guessing.  The advantage of returning the whole object is it will let me use whatever 'goodness' measure I after the fact - even use several, to determine which model I want to keep.

## purrr

`purrr` is a package from the tidyverse that provides a set of tools to improve functional programming.  It has consistent syntax versions of `map` and `reduce` type functions that make code more readable than the base versions in R.  But you'll notice I used `lapply` above instead of `purrr::map`. I'm used to using `lapply` and I don't like loading a whole package for simple use cases like that.  Where I do like using it is for more complex cases - like `at_depth`

### at_depth

`lapply` takes an object and applies a function to each element of the object and returns it in a list.  Consider a heirarchical list like the one below.

```{r}
dumbList = list(z=list(a=c(1,2), b=c(2,3)), y=list(a=c(1,1),b=c(1,4)), x=list(a=c(2,2),b=c(2,0)))
```

Maybe I want to sum each element so that z$a = 3, y$a = 2, x$a = 4, etc...

```{r,error=TRUE}
lapply(dumbList,sum)
```

It doesn't work because `dumbList[[1]]` is a list and you can't sum a list.  However, with `purrr::at_depth` we can specify the depth we want to work at - in this case, the second level.  Operate on the lists inside of lists.

```{r}
library(purrr) #version 0.2.2
purrr::at_depth(dumbList,2,sum)
```

This is equivalent to the following code which requires an anonymous function to work.

```{r,eval=FALSE}
lapply(dumbList,function(x){
  lapply(x,sum)
})
```

How does this relate to `allSubsetsRegression`?  Let's look at the results of running it.

### Results of allSubetsRegression

The function returns a hierarchical list - the top level has $k$ elements and the second level has $i$ choose $k$ elements.

```{r}
res = allSubsetsRegression(df,"Y")
length(res) #4 because we had 4 independent variables
lapply(res,length) # 4, 6, 4, 1 because there are 4 ways to select 1 variable from a group of 4, 6 ways to select 2, 4 ways to select 3 and 1 way to select 4
```

This allows me to easily index into any specific model via `res[[k]][["terms"]]` and inspect that model.  Want to see `X1+X2+X3`?

```{r}
res[[3]][["X1+X2+X3"]]
```

Now I want to calculate goodness statistics on each of these so I can determine the best models.

### regGoodness

`regGoodness` (regression goodness) operates on the hierarchical list produced by `allSubsetsRegression`, applying a function which produces a 'goodness' score and then sorts by if that goodness score should be minimzed or maximized.  It does this by using `purrr::at_depth` to allow a clean hierarchical output when generating the models, but ease of computing summary statistics afterward.  **note** this does leverage the `stringr` package.

```{r}
library(stringr) #version 1.2.0

regGoodness = function(models,f,direction){
  if(!(direction %in% c(">","<"))) stop("Invalid direction - must be < or >")
  if(!is.function(f)) stop("f must be a function")
  if(!("subReg" %in% class(models))) stop("models must be generated by allSubsetsRegression")
  score = unlist(purrr::at_depth(models,2,f))
  output = data.frame(N = stringr::str_count(names(score),"\\+")+1,Terms = names(score),Value = score,stringsAsFactors = FALSE)
  rownames(output) = NULL
  if(direction == "<"){
    return(output[order(output$Value),])
  } else {
    return(output[order(output$Value,decreasing = T),])
  }
}
```

For example, we could get the $R^2_{Adj}$ for each model by using this function.

```{r}
asrAR2 = function(model){
  summary(model)$adj.r.squared
}
```

We pass it into the function thusly and get out a data.frame with summary details about each model fit.

```{r}
ar2 = regGoodness(res,asrAR2,">")
ar2
```

### Coefficient Stability

We can also analyze how stable our coefficient estimates are as we vary what is in or out of the model.  First we need to extract the coefficients from each model, then mutate the vector so it becomes a named data frame.  The `dplyr::bind_rows` function is great because, like a "UNION ALL" query in SQL, keeps all the column names as it binds data frames together like new rows.

```{r}
library(dplyr) #version 0.5.0
coefs = purrr::at_depth(res,2,coefficients) %>% 
  purrr::at_depth(2,function(x){as.data.frame(t(as.matrix(x)))}) %>% 
  purrr::reduce(bind_rows) %>% 
  mutate(Terms = ar2$Terms)
coefs
```

## Conclusion

All subsets regression presents many challenges, theoretical and practical.  I was able to easily write some functional code which produced a hierarchical list.  The structure is nice for exploring the results but presents hurdles when trying to analyze it with functional programming.  `purrr::at_depth` allows for easy functional programming on hierarchical lists without the need for anonymous functions or other cumbersome overhead.