---
title: "mgsub v1.0 Launched to CRAN"
author: "Mark Ewing"
description: "Multiple, simultaneous string substitions done safely."
date: 2018-01-26
categories: ["r"]
type: "post"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(htmlTable)
library(tableHTML)
prepTable = function(r,t){
  dt = data.frame(Library=names(r),
                  Result =unlist(r,use.names = FALSE),
                  Correct=unlist(r,use.names = FALSE) == t,
                  stringsAsFactors = FALSE)
  return(dt)
}
```

## Official CRAN Launch

Earlier this week I submitted `mgsub` to CRAN and after a couple of days [it was accepted](https://cran.r-project.org/web/packages/mgsub/index.html)! Now it's live!  I'm very excited to have published my second package and one that I think is a more valuable contribution than my first.  The package represented a few firsts for me. The first package that I wrote tests for, checked code coverage on and for which I wrote a vignette.  Woot!

In my nervous anticipation after submission I worried I might have missed the email indicating it went live so I did a quick search for `mgsub` to see what would come back.  Turns out, there are at least 4 other implementations of `mgsub` in packages on CRAN already.  Nothing kicks imposter syndrome into overdrive like seeing that 4 other people already did what you did.  But the focus of my implementation is *safety* - it's important to know that a string manipulation function is going to work the way you intend.  If your processing enough strings (or even a big enough string) it is difficult to do QA on a function.  Spot checking could miss things and if the process isn't safe enough, there may be no way of accomplishing your goal.

So, I decided to download all 4 implementations and test them to see how they stack up.  Get ready for lots of examples!

## Contenders

The 4 contenders are `qdap` (which I covered in a [previous post](http://thug-r.life/post/2018-01-10-safe-multiple-string-substitutions/)), `bayesbio` (which actually doesn't export their function), `bazar` and `textclean`.  Note that I'm not actually loading the libraries because they will all have a namespace collision on `mgsub` so I'm calling by reference.

```{r}
# library(mgsub)
# library(qdap)
# library(bayesbio) #not exported
# library(bazar)
# library(textclean)
```

To make testing/comparing easier I wrote a function for each package which accepts a list of the string to be modified, the vector of matches and the vector of replacements.  This way I can put the functions in a list and just `lapply` (the name of `f` for my list is super expressive).  One thing to note is that I have ensured that each function accepts regular expression input (by setting `fixed=FALSE` where necessary).

```{r}
mgsub = function(a){
  replace = a$replace
  names(replace) = a$match
  mgsub::mgsub(a$orig,replace)
}
qdap = function(a){
  qdap::mgsub(a$match,a$replace,a$orig,fixed=FALSE)
}
bayesbio = function(a){
  bayesbio:::mgsub(a$match,a$replace,a$orig)
}
bazar = function(a){
  bazar::mgsub(a$match,a$replace,a$orig)
}
textclean = function(a){
  textclean::mgsub(a$orig,a$match,a$replace,fixed = FALSE)
}
f = list(mgsub,qdap,bayesbio,bazar,textclean)
names(f) = c("mgsub","qdap","bayesbio","bazar","textclean")
```

Each test will cover how well each package handles different scenarios in multiple, global string substition.  For each test I provide the original string, the matches, the replacements *and* the target result.  Each test will return a table which contains the result of the call and indicates how successful each library was.

## Simple Test

We'll start off nice and easy.  We'll modify a string by replacing "hey" with "yo" and "let's" with "we".

```{r}
a = list(
orig = "hey ho, let's go!",
match = c("hey","let's"),
replace = c("yo","we")
)
simple_target = "yo ho, we go!"
simple = lapply(f,function(x) x(a))

htmlTable(prepTable(simple,simple_target),
          align="llr",
          css.cell = "padding-left: 1em; padding-right: 1em")
```

Right away we see that `bayesbio` is doing something very different from expected. Spoiler Alert: it continues. Given that it's not an exported function, it may have a very specific use case in mind.  Otherwise, the other packages all make the correct substitutions.

## Substring

Next we'll test how the libraries protect against substring substitution. In this case we want to replace "the" with "any" and "they" with "we".  I specifically put "the" earlier in the list of matches.  If the functions aren't safe, they may detect "the" as a substring of "they" and the result would be "any".  What's very problematic about this example is that "any" is a real word which would make detection even more difficult when scanning results.

```{r}
a = list(
orig = "they don't know the answer",
match = c("the","they"),
replace = c("an","we")
)
substring_target = "we don't know an answer"
substring = lapply(f,function(x) x(a))

htmlTable(prepTable(substring,substring_target),
          align="llr",
          css.cell = "padding-left: 1em; padding-right: 1em")
```

The other packages have made the substring mistake.  Now, this was an engineered example - by providing "the" before "they", anything that just matches in order will have this problem.  It could be avoided by the user simply sorting their matches by `nchar`.  `qdap` actually does this by default - *if* `fixed = TRUE` (the actual argument is `order.pattern` but it seems to ignore setting it to `TRUE` if `fixed = FALSE`).  So, just know that if you're only working with fixed matches, `qdap` would have worked correctly here.

```{r}
qdap::mgsub(c("the","they"),c("an","we"),"they don't know the answer",fixed=TRUE)
```

## Transpose words

Transposing words means taking a pair of words and replacing each with the other.  Here, we replace "hey" with "ho" and "ho" with "hey".

```{r}
a = list(
orig = "hey ho, let's go!",
match = c("hey","ho"),
replace = c("ho","hey")
)
transpose_target = "ho hey, let's go!"
transpose = lapply(f,function(x) x(a))

htmlTable(prepTable(transpose,transpose_target),
          align="llr",
          css.cell = "padding-left: 1em; padding-right: 1em")
```

The failure is amongst all the contenders again.  Note, in this case, `qdap` fails even if `fixed = TRUE`. The problem (at least for `bazar` and `textclean`) is that they're simply looping through matches. "hey" is replaced with "ho" leaving "ho ho, let's go" and then "ho" is replaced with "hey" generating the result.

```{r}
qdap::mgsub(c("hey","ho"),c("ho","hey"),"hey ho, let's go!",fixed=TRUE)
```

## Shifting Words

This is similar to shifting words except it's a chain of shifts.  By shifting each word one spot to the left (with wraparound) we check if it's a simple looping problem or something else.  Also note, each match here is the same number of characters so any placeholder work done based on `nchar` would fail.

```{r}
a = list(
orig = "hey, how are you?",
match = c("hey","how","are","you"),
replace = c("how","are","you","hey")
)
shift_target = "how, are you hey?"
shift = lapply(f,function(x) x(a))

htmlTable(prepTable(shift,shift_target),
          align="llr",
          css.cell = "padding-left: 1em; padding-right: 1em")
```
In this case, every word has been replaced with the last replacement value (again, due to the looping).

## Regex

Next I test regular expression support (including backreferences.)

```{r}
a = list(
orig = "Dopazamine is not the same as dopachloride or dopamezamine and is still fake.",
match = c("[Dd]opa([^ ]*?mine)","fake"),
replace = c("Meta\\1","real")
)
regex_target = "Metazamine is not the same as dopachloride or Metamezamine and is still real."
regex = lapply(f,function(x) x(a))

htmlTable(prepTable(regex,regex_target),
          align="llr",
          css.cell = "padding-left: 1em; padding-right: 1em")
```

This one passes easily (everything is working with a form of `gsub` turned on) though remember we did have to explicitly flag `fixed = FALSE` for `qdap` and `textclean`.  Forgetting to do so would have resulted in failure to match or weird replacements.

## Regex Susbtring

Most things failed to protect against substring matches, but what about when the smaller string (in terms of `nchar`) is a regular expression that matches a variable number of characters?  Does the function actively determine what is a substring and what is a longer string?  The example below is *super* engineered to create the scenario.

```{r}
a = list(
orig = "Dopazamine is a fake chemical",
match = c("Dopazamin","Do.*ne"),
replace = c("freakout","metazamine")
)
regex_substring_target = "metazamine is a fake chemical"
regex_substring = lapply(f,function(x) x(a))

htmlTable(prepTable(regex_substring,regex_substring_target),
          align="llr",
          css.cell = "padding-left: 1em; padding-right: 1em")
```

Note that only `mgsub::mgsub` correctly protects substrings in the presence of variable length regular expressions - even when that variable length is presented later in the list of inputs.

## Speed

So I've shown several examples of cases where other implementations of `mgsub` fail to safely perform substitutions. But what is the performance cost?

### Simple

```{r}
library(microbenchmark)

s = "Hi, my name is Mark"
m = c("Hi","Mark")
r = c("Goodbye","Tom")
names(r) = m

smb = microbenchmark(
  mgsub = mgsub::mgsub(s,r,fixed=TRUE),
  qdap = qdap::mgsub(m,r,s),
  bayesbio = bayesbio:::mgsub(m,r,s),
  bazar = bazar::mgsub(m,r,s),
  textclean = textclean::mgsub(s,m,r)
)
smb = print(smb)
```

`mgsub` and `qdap` in this simple case are in the same order of magnitude.  The other three are significantly faster - the slowest of them is still 2x faster than `mgsub` or `qdap`.

### Regex

```{r}
s = "Dopazamine is not the same as Dopachloride and is still fake."
m = c("[Dd]opa(.*?mine)","fake")
r = c("Meta\\1","real")
names(r) = m

rmb = microbenchmark(
  mgsub = mgsub::mgsub(s,r),
  qdap = qdap::mgsub(m,r,s,fixed=FALSE),
  bayesbio = bayesbio:::mgsub(m,r,s),
  bazar = bazar::mgsub(m,r,s),
  textclean = textclean::mgsub(s,m,r,fixed=FALSE)
)
rmb = print(rmb)
```

When we add regular expressions (with backreferences) the speed difference grows even bigger.

All that being said, things are still in microseconds, so it's not necessarily a practical difference.  And, is it worth being so much faster knowing you could be getting wrong results?

## Conclusion

```{r,warning=FALSE,echo=F}
final = dplyr::tibble(
  `Library&nbsp;&nbsp;` = names(f),
  `&nbsp;&nbsp;Simple&nbsp;&nbsp;` = ifelse(unlist(simple) == simple_target,"Y","N"),
  `&nbsp;&nbsp;Substring&nbsp;&nbsp;` = ifelse(unlist(substring) == substring_target,"Y","N"),
  `&nbsp;&nbsp;Transpose&nbsp;&nbsp;` = ifelse(unlist(transpose) == transpose_target,"Y","N"),
  `&nbsp;&nbsp;Shift&nbsp;&nbsp;` = ifelse(unlist(shift) == shift_target,"Y","N"),
  `&nbsp;&nbsp;Regex&nbsp;&nbsp;` = ifelse(unlist(regex) == regex_target,"Y","N"),
  `Regex<br>Substring` = ifelse(unlist(regex_substring) == regex_substring_target,"Y","N"),
  `Simple ` = round(smb$median,0),
  `Regex ` = round(rmb$median,0)
)
rownames(final) = names(f)
# htmlTable(final,
#           cgroup = c("Correctness Results","Median Runtime (microseconds)"),
#           n.cgroup = c(6,2),
#           align="llllllrr",
#           css.cell = "padding-left: 1em; padding-right: 1em")
tableHTML(final,border = 0,rownames=FALSE,escape=FALSE,collapse="collapse",
          second_header = list(c(1,6,2),c("","Correctness Results","Median Runtime (microseconds)"))) %>% 
  add_css_row(css = list("border-top","2px solid black"),rows=1) %>% 
  add_css_row(css = list("border-bottom","1px solid black"),rows=2) %>% 
  add_css_second_header(css = list("border-bottom","1px solid black"), second_headers=c(2,3)) %>% 
  add_css_header(css = list("text-align","right"), headers = c(8,9)) %>% 
  add_css_row(css = list("border-bottom","2px solid black"),rows=7) %>% 
  add_css_column(css = list("text-align","center"),columns = c(2:7)) %>% 
  add_css_column(css = list("text-align","right"),columns = c(8,9)) %>% 
  add_css_conditional_column(conditional = '==',
                             value = "Y",
                             css = list('background-color','#8df086'),
                             columns = 1:9) %>% 
  add_css_conditional_column(conditional = '==',
                             value = "N",
                             css = list('background-color','#6f87fc'),
                             columns = 1:9)
```

I've shown that only `mgsub::mgsub` provides actual correct multiple, global string substitution from the 5 packages that have a function named `mgsub` in every possible scenario. While it is on the slower end of runtime, it's a small price to pay for safety.

