---
title: "How many random numbers does it take?"
author: "Mark Ewing"
description: "Simulating our way into fun math outcomes."
date: 2017-10-30
categories: ["r"]
type: "post"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Fermat and his library

This morning I woke up to a delightful tweet from fermatslibrary about sample random uniform numbers and how many it takes, on average, to sum to 1.

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Pick a uniformly random number in [0,1] and repeat until the sum of the numbers picked is &gt;1. You&#39;ll on average pick e≈2.718... numbers! <a href="https://t.co/8ak1hYENCi">pic.twitter.com/8ak1hYENCi</a></p>&mdash; Fermat&#39;s Library (@fermatslibrary) <a href="https://twitter.com/fermatslibrary/status/924263998589145090?ref_src=twsrc%5Etfw">October 28, 2017</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

If you look at the embedded picture, you can see the math sketched out but of course it's alwasy more fun to simulate.

## Some R Code

How can we generate a random uniform [0,1] number in R?  There are many distributions built in and this one is included.  Note, I'm setting a seed to ensure if you're following along at home you will get the same answer.  `runif` is the function which generates random values, by default *min* is 0 and *max* is 1 (just what we want).

```{r}
set.seed(1001)
runif(n = 1,min = 0,max = 1)
```

We can visually 'verify' the distribution we're drawing from with a histogram.  While it's not totally flat, it's fairly obvious that it's uniformly sampling values between 0 and 1.

```{r}
hist(runif(10000), main = "Histogram of 10,000 Random Uniform [0,1] Values",xlab="")
```

### How many are needed?

To determine how many values are needed to sum to 1, we'll use a `while` loop.  This is a loop that will cycle until a logical condition is satisified.  It's like a `for` loop except it has the potential to be an infinite loop.

```{r}
summedValue = 0
steps = 0
while(summedValue < 1){
  summedValue = summedValue + runif(1)
  steps = steps+1
}
cat("Total Summed Value:",summedValue,"\n")
cat("It took ",steps," step",ifelse(steps==1,"","s")," to get there\n",sep="")
```

This is great! We can see to total value, and how many values were drawn to get to 1.  But now we need to do this lots of times.

### Functional programming

By wrapping our above code in a function we'll be able to take advantage of functional programming.  Note how the function is mostly a wrapper and all that's changed is we replaced the `cat` steps with a `return` argument (what the function will return as a value).

```{r}
countUnif = function(){
  summedValue = 0
  steps = 0
  while(summedValue < 1){
    summedValue = summedValue+runif(1)
    steps = steps+1
  }
  return(steps)
}

countUnif()
```

Now we can simulate this many, many times. The `replicate` function makes it very easy to generate 1,000,000 experiments.

```{r}
results = replicate(1000000,countUnif())
cat("Average values required from simulation:",mean(results),"\n")
cat("Average values required from math      :",exp(1),"\n")
```

Wow! That's really close.  But I'd really like to see how the value stabilizes over time.  So, I'd like a cumulative average.

```{r}
cumulativeAverage = cumsum(results)/(1:1000000)
```

## Visualize!

```{r}
plot(cumulativeAverage,type='l',main='Average Random Uniform [0,1] Values Needed to Sum to 1',ylab='Cumulative Average')
abline(h=exp(1),col='red')
```

Zoom and enhance!  Here we can see that we're not necessarily stabilizing,but the movement is all in a very small range.

```{r}
plot(cumulativeAverage,type='l',main='Average Random Uniform [0,1] Values Needed to Sum to 1',ylab='Cumulative Average',ylim=c(2.715,2.72))
abline(h=exp(1),col='red')
```

## Conclusion

A fun and easy simluation to do in R, allowing functional programming practice, to demonstrate a neat math fundamental.