---
title: "Safe, Multiple String Substitutions with mgsub::mgsub"
author: "Mark Ewing"
description: "Solving an infrequent problem with a package."
date: 2018-01-10
type: "post"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## String substitutions

Note - the package I wrote was originally inspired by a challenge a coworker tossed out. It also happened to provide a solution to [this SO question](https://stackoverflow.com/questions/48169135/str-replace-all-replacing-named-vector-elements-iteratively-not-all-at-once/48189012#48189012) which was really cool!

Substitutions in strings are best handled with *regular expressions* which are an amazingly powerful and flexible tool.  Regular expressions are a way of expressing patterns in strings.  In the example below I want to find the four letters, "dopa" and replace them with "meta". I accomplish this with the internal function `sub`.

```{r}
sub("dopa","meta","The chemical dopaziamine is fake, long live dopaziamine!")
```

## Multiple substitutions

Sometimes you need to substitute many things, all at once.

### Multiple instances of the same pattern

In the first example, only the first instance of "dopa" was replaced with "meta".  If the goal is to replace all of them, I can use the internal function `gsub` (the *g* stand for global!).

```{r}
gsub("dopa","meta","The chemical dopaziamine is fake, long live dopaziamine!")
```

However, the internal regex functions aren't vectorized, so I can't have multiple patterns.

```{r}
gsub(c("dopa","fake"),c("meta","real"),"The chemical dopaziamine is fake, long live dopaziamine!")
```

### Multiple patterns

There are a few string substitution methods that handle multiple patterns.

#### stringr

The function `str_replace_all` in the `stringr` package supports vectorized patterns and replacements. However, it applies each set individually giving `n` responses (where `n` is the longer length of the pattern or replacement vector).  So, this doesn't really work.

```{r}
stringr::str_replace_all("The chemical dopaziamine is fake, long live dopaziamine!"
                         ,c("dopa","fake"),c("meta","real"))
```

#### qdap

The function `mgsub` in the `qdap` package also supports vectorized patterns and replacements.  It works by using placeholders and then iteratively applying internal regex functions.

```{r}
qdap::mgsub(c("dopa","fake"),c("meta","real"),"The chemical dopaziamine is fake, long live dopaziamine!")
```

#### chartr

There's a special case which is mostly for transliteration, so it only works on single characters.  `chartr`, an internal function, takes a string of old characters and a string of new characters and does simultaneous replacement, characterwise, on the string.  This does not support regex or anything other than single characters, so it's pretty limited.

```{r}
chartr("ho","oh","ho ho hoot")
```

### Problems with safety

I noted that `qdap::mgsub` uses placeholders.  This can actually cause a problem in certain cases where patterns are the same length.  Consider the phrase "Hey, how are you?" where I want to shift each word to the left. So, "hey" shoudl be replaced by "how", "how" by "are", etc.  Note, each pattern to match is 3 characters long.  Note also that I pass in the `ignore.case=T` argument so my patterns won't be bothered by capitalization.

```{r}
qdap::mgsub(c("hey", "how", "are", "you"),
            c("how", "are", "you", "hey"),
            "Hey, how are you?",
            fixed=F,ignore.case=T)
```

The placeholders became indistinguishable and so every word was replaced with the same word.

### A safer option

I just published a [github repo](https://github.com/bmewing/mgsub) which contains a new R package called `mgsub`.  It is a safe alternative to `qdap::mgsub`, fully supporting regular expression matching and replacement in a way that guarantees safety.  It also replaces with the longer match first, so sub-matches won't mess things up.

Also, the code is pure R (for now) with no dependencies, so you won't get a lot of bloat.

#### Installing from github

```{r,eval=F}
devtools::install_github("bmewing/mgsub")
```

#### Examples

First, the case that broke `qdap`.

```{r}
mgsub::mgsub("Hey, how are you?", c("hey", "how", "are", "you"), c("how", "are", "you", "hey"),
             ignore.case=T)
```

We can also try a complex regular expression.  Note we use regular expressions in the match and the replacement and it works exactly as expected.  We only replaced "dopa" with "meta" with it's a -mine group as opposed to a -ride group.  Disclaimer - I know nothing about chemistry and stuff, so I don't know if those are real.

```{r}
mgsub::mgsub("Dopazamine is not the same as Dopachloride and is still fake.",
             c("[Dd]opa(.*?mine)", "fake"),
             c("Meta\\1", "real"),
             ignore.case=F)
```

We can see the substring protection here. Even though "the" is a substring of "they" and appears in the list first, "they" is given priority when it is also matched.

```{r}
mgsub::mgsub("They don't understand the value of what they seek.",
             c("the", "they"),
             c("a","we"),
             ignore.case=T)
```

You can also use it on single characters.

```{r}
mgsub::mgsub("ho ho hoot", c("h", "o"), c("o", "h"))
```

### Development plans

Once I write unit tests and get some error handling in place I will submit to CRAN.  Then I'll start working on getting the code into C++ to test the performance enhancements.  Overall goal will be *low* overhead.